---
title: "Total Cost Analysi"
author: "Mike Gaunt"
date: "`r Sys.Date()`"
output: github_document
knit: (function(input, ...) {
    rmarkdown::render(
      input,
      output_file = paste0(
        xfun::sans_ext(input), '-', Sys.Date(), '.html'
      ),
      envir = globalenv()
    )
  })
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  cache = FALSE, cache.lazy = FALSE, autodep = TRUE, warning = FALSE, 
  message = FALSE, echo = TRUE, dpi = 180,
  fig.width = 8, fig.height = 5, echo = FALSE
  )

```

```{r include=FALSE}
#sources local global file
source(here::here("code/zz_localG.r")) 

#loads the data
# o_wo = readRDS(here("data/o_wo.rds"))
# o_wo_agg = readRDS(here("data/o_wo_agg.rds"))
# udcm = readRDS(here("data/udcm.rds"))
joined = readRDS(here("data/joined_2021-05-03.rds"))
```

```{r}
veh_costs = read_excel(
  here("data/King_County_Metro.Asset_Inventory.2020_20210503.xlsx"),
  sheet = "Rolling stock form_mg",
  skip = 10) %>%
  clean_names() %>% 
  data.table() %>%  
  .[,`:=`(make_model = make_model %>%  str_to_lower())]

veh_costs_agg = veh_costs %>%  
  .[,.(ave_cost = mean(replacement_cost, na.rm = T),
       med_cost = median(replacement_cost, na.rm = T)), by = make_model] %>%  
  separate(col = "make_model", into = c("make", "model"), sep = "/") %>%  
  filter(!is.na(model))

working_list = joined %>%
  .[, .(manufacturer, make, model)] %>%  
  .[order(manufacturer, model)] %>% 
  unique() %>%  
  mutate(across(c('manufacturer', 'make', 'model'), str_to_lower)) 

```

# Summary 

This analysis was a quicky - I wanted to see if age had anything to do with the total cost curves. I had presumed that mileage and age were highly correlated and that total cost (maintance + replacement) would be practically the same when compared against a vehicle's mileage or age. This presumption was based off of very similar maintance cost versus age and mileage curves seen in the maintenance analysis. 

But after some further investigation, it appears that the age of a vehicle does in fact create a different total cost curve. 
 
# The analysis


```{r fig.height=12, fig.width=8}
yolo = joined %>%  
  .[order(unit_id, fisc_pd_corr)] %>% 
  #making running value for maintenance by units
  .[,`:=`(running_cost = cumsum(event_cost_max)), by = .(unit_id)] %>% 
  .[,`:=`(running_cc = running_cost/meter_max_mnth)] %>%  
  #removes cost values cause by poor data - no miles driven % negative cost
  filter(!is.infinite(event_cost_per_meter_per_mnth) &
           event_cost_per_meter_per_mnth >= 0) %>% 
  mutate(across(c('manufacturer', 'make', 'model'), str_to_lower)) %>% 
  merge.data.table(veh_costs_agg, by = c("model", "make")) %>% 
  #creates some new cost values
  .[,`:=`(rep_cost = ave_cost/meter_max_mnth)] %>% 
  .[,`:=`(comb_cost = rep_cost + running_cc)] %>% 
  mutate(`Make/Model` = paste0(make, "\n", model) %>%  
           str_to_title() %>%
           str_replace_all("Ii", "II")) %>% 
  .[,.(median = median(comb_cost)), by = .(`Make/Model`, bin = as.numeric(floor_divide(meter_max_mnth, 1000)))] %>%
  .[order(`Make/Model`, bin)]  %>% 
  .[,`:=`(median_roll = frollmean(median, n = 5, align = "center")), by = .(`Make/Model`)] %>%
  pivot_longer(cols = c(median, median_roll)) %>% 
   filter(`Make/Model` != "Chevrolet\nUplander" #&
         ) %>% 
      ggplot() + 
  geom_line(aes(bin, value, color = name), size = 1) + 
  coord_cartesian(ylim=c(0, 20)) +
  facet_grid(rows = vars(`Make/Model`)) +
  theme() + 
  labs(y = "Costs per Unit of Operation", 
       x = "Vehicle Odometer", 
       color = "Total Cost of Operation", 
       caption = "Vertical plot range limited [0, 20] for visual clarity." ) + 
  theme(axis.text.x = element_text(angle = 45, vjust = 0.5, hjust=.5), legend.position = "top") +
  scale_color_discrete(
    labels = c("W/o Rolling Window", "W/ Rolling Window")
  )
```

```{r}
# #najes median values by meter bins 
# data2 = data1 %>% 
#   group_by(meter_bin, `Make/Model`, name) %>%  
#   summarise(median = median(value, na.rm = T), 
#             sd = sd(value, na.rm = T),
#             sqrtN = sqrt(n()),
#             semd = 1.253*sd/sqrtN) %>%  
#   mutate(meter_bin = as.numeric(as.character(meter_bin))) 
# 
# #makes median values by age bins
# data2_age = data1 %>% 
#   filter(!is.infinite(value)) %>% 
#   group_by(age_bin, `Make/Model`, name) %>%  
#   summarise(median = median(value, na.rm = T), 
#             sd = sd(value, na.rm = T),
#             sqrtN = sqrt(n()),
#             semd = 1.253*sd/sqrtN) %>%  
#   mutate(age_bin = as.numeric(as.character(age_bin))) 
```

## Box/Line Plots {.tabset}

### Meter Bin
```{r fig.width = 10, fig.height = 12}
# data2 %>% 
#   filter(name == "Total Cost") %>% 
#   ggplot() +
#   coord_cartesian(ylim=c(0, 2)) +
#   geom_boxplot(data = data1 %>%
#                  filter(name != "Total Cost"), 
#                aes(meter_max_mnth, value, color = name, group = interaction(name, meterbinnn, lex.order = F)),
#                outlier.shape = NA) +
#   geom_point(aes(meter_bin, median, color = name)) +
#   geom_line(aes(meter_bin, median, color = name))  +
#   # facet_grid(rows = vars(`Make/Model`), scales = "free_x") +
#   facet_wrap(vars(`Make/Model`), ncol = 2) +
#   theme(axis.text.x = element_text(angle = 45, vjust = 0.5, hjust=.5)) + 
#   labs(y = "Costs per Unit of Operating Time", 
#        x = "Vehicle Mileage (20K mile bins)", 
#        color = "Cost Types", 
#        # title = "Vehicle Operation Costs per Unit of Operation Time",
#        caption = "Vertical plot range limited [0, 2] and outliers omitted for visual clarity." ) + 
#   theme(legend.position="top")
```

### Age Bin
```{r fig.width = 12, fig.height = 12, eval=FALSE}
data2_age %>% 
  filter(name == "Total Cost") %>% 
  ggplot() +
  coord_cartesian(ylim=c(0, 2)) +
  geom_boxplot(data = data1 %>%
                 filter(name != "Total Cost"), 
               aes(age_bin, value, color = name, group = interaction(name, age_bin, lex.order = F)),
               outlier.shape = NA) +
  geom_point(aes(age_bin, median, color = name)) +
  geom_line(aes(age_bin, median, color = name))  +
  facet_grid(rows = vars(`Make/Model`), scales = "free_x") +
  theme(axis.text.x = element_text(angle = 45, vjust = 0.5, hjust=.5)) + 
  labs(y = "Costs per Unit of Operating Time", 
       x = "Vehicle Age (years since mfg.)", 
       color = "Cost Types", 
       title = "Vehicle Operation Costs per Unit of Operation Time",
       caption = "Vertical plot range limited [0, 2] and outliers omitted for visual clarity." ) + 
  theme(legend.position="top")
```

## {-}

## Line Plots {.tabset}

### Meter Bin
```{r fig.width = 12, fig.height = 12, eval=FALSE}
data2 %>% 
  ggplot() +
  coord_cartesian(ylim=c(0, 2)) +
  # geom_ribbon(aes(age_bin, ymin = median-semd, ymax = median+semd, fill = name), alpha = .3) +
  geom_point(aes(meter_bin, median, color = name)) +
  geom_line(aes(meter_bin, median, color = name)) +
  facet_grid(rows = vars(`Make/Model`), scales = "free_x") +
  theme(axis.text.x = element_text(angle = 45, vjust = 0.5, hjust=.5)) + 
  labs(y = "Costs per Unit of Operating Time", 
       x = "Vehicle Age (years since mfg.)", 
       color = "Cost Types", 
       title = "Vehicle Operation Costs per Unit of Operation Time",
       caption = "Vertical plot range limited [0, 2] and outliers omitted for visual clarity." ) + 
  theme(legend.position="top")
```
### Age Bin
```{r fig.width = 12, fig.height = 12, eval=FALSE}
data2_age %>% 
  ggplot() +
  coord_cartesian(ylim=c(0, 2)) +
  # geom_ribbon(aes(age_bin, ymin = median-semd, ymax = median+semd, fill = name), alpha = .3) +
  geom_point(aes(age_bin, median, color = name)) +
  geom_line(aes(age_bin, median, color = name)) +
  facet_grid(rows = vars(`Make/Model`), scales = "free_x") +
  theme(axis.text.x = element_text(angle = 45, vjust = 0.5, hjust=.5)) + 
  labs(y = "Costs per Unit of Operating Time", 
       x = "Vehicle Age (years since mfg.)", 
       color = "Cost Types", 
       title = "Vehicle Operation Costs per Unit of Operation Time",
       caption = "Vertical plot range limited [0, 2] and outliers omitted for visual clarity." ) + 
  theme(legend.position="top") 
```
## {-}


Key takeaways (20210504): 

* Key takeways from the initial iteration of this analysis still stand
  + Both replacement and maintenacne costs either become asymptotic or slowly dirft downward or upwards, respectively 
  + No definitive trough behaviopur seen in combined cost values
    - Initial assumption: combined costs would bend upwards at some point given pressure from increasing maintenace costs
  + Vehicles do exhibit an increase in combined cost at extreme ranges 
    - Likely, due to a small number of maintance events at these range bins 
  + A few vehicles show a slight but noticable increase in comvined costs
    - Uplander ~ 150K
    - Entervan ~ 150K
    - Express ~ 150K
    - GC 2 ~ 225K


```{r eval=FALSE}
data2 %>% 
  filter(str_detect(`Make/Model`, "Sup")) %>% 
  mutate(name = fct_relevel(name, c("Corrective Cost", "Preventative Cost", "Total Cost"))) %>% 
  ggplot() +
  coord_cartesian(ylim=c(0, 2)) +
  geom_point(aes(meter_bin, median, color = name)) +
  geom_line(aes(meter_bin, median, color = name)) +
  facet_grid(rows = vars(`Make/Model`), scales = "free_x") +
  theme(axis.text.x = element_text(angle = 45, vjust = 0.5, hjust=.5)) + 
  labs(y = "Preventative Costs", 
       x = "Vehicle Mileage (20K mile bins)", 
       color = "Cost Types", 
       # title = "Vehicle Operation Costs per Unit of Operation Time",
       caption = "Vertical plot range limited [0, 2] for visual clarity." ) + 
  theme(legend.position="top") 
```
