---
title: "Combined Costs Analysis"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  cache = FALSE, cache.lazy = FALSE, autodep = TRUE, warning = FALSE, 
  message = FALSE, echo = TRUE, dpi = 180,
  fig.width = 8, fig.height = 5, echo = FALSE
  )
```

```{r include=FALSE}
#sources local global file
source(here::here("code/zz_localG.r")) 

#loads the data
# o_wo = readRDS(here("data/o_wo.rds"))
# o_wo_agg = readRDS(here("data/o_wo_agg.rds"))
# udcm = readRDS(here("data/udcm.rds"))
joined = readRDS(here("data/joined_2021-05-03.rds"))
```

```{r}
veh_costs = read_excel(
  here("data/King_County_Metro.Asset_Inventory.2020_20210503.xlsx"),
  sheet = "Rolling stock form_mg",
  skip = 10) %>%
  clean_names() %>% 
  data.table() %>%  
  .[,`:=`(make_model = make_model %>%  str_to_lower())]

veh_costs_agg = veh_costs %>%  
  .[,.(ave_cost = mean(replacement_cost, na.rm = T),
       med_cost = median(replacement_cost, na.rm = T)), by = make_model] %>%  
  separate(col = "make_model", into = c("make", "model"), sep = "/") %>%  
  filter(!is.na(model))

working_list = joined %>%
  .[, .(manufacturer, make, model)] %>%  
  .[order(manufacturer, model)] %>% 
  unique() %>%  
  mutate(across(c('manufacturer', 'make', 'model'), str_to_lower)) 

```

# Summary 

This analysis attempts to determine the most cost efficient time to remove a vehicle from the KCM fleet given vehicle type and maintenance costs.

20210503 Changes:
Data was manually edited in both the joined and vehicle cost datasets. More specificlaly, vehicle manufacture/make/model informating were edited to be consistent acorss both data sets, in turn, increasing the number and quality of merges between both data sets on said values.

Performing this process resulted in more data included in the combiuned cost analysis due to the decrease in failed merges which had been produced from spurious typos in the merged data fields.

# The Data

The data used for this analysis consists of the KCM work order data used in the previous maintenance cost analysis and vehicle cost data as supplied by NR. The vehicle cost data was imperfect data and required some cleaning such that it could be combined with he maintenance cost data set. The vehicle cost data can be seen below:

```{r}
veh_costs_agg %>%  
  dt_common()
```

This table was joined with the KCM work order data, the join was performed on vehicle model type. The table below details unique vehicle types and their average and median replacement costs resulting from the inner join:
```{r}
working_list %>%  
  merge.data.table(veh_costs_agg, by = c("model", "make")) %>%  
  dt_common()
```
 
# The analysis

For this analysis, replacement and operational costs were aggregated (average/median) per meter bin per unique model type. Those two metrics were superimposed to determine if there is a consistent point in a vehicle's life when it becomes too expensive to maintain, i.e. maintenance costs overcome replacement  costs. 

The maintenance and replacement costs given meter mileage for a number of vehicle types can be seen in the plot below. 

```{r fig.width = 12, fig.height = 12}
data1 = joined %>%  
  mutate(across(c('manufacturer', 'make', 'model'), str_to_lower)) %>% 
  merge.data.table(veh_costs_agg, by = c("model", "make")) %>% 
  .[,`:=`(rep_cost = ave_cost/meter_max_mnth)] %>% 
  .[,`:=`(comb_cost = rep_cost + event_cost_per_meter_per_mnth)] %>% 
  pivot_longer(cols = c(rep_cost, event_cost_per_meter_per_mnth, comb_cost)) %>% 
  mutate(`Make/Model` = paste0(make, "\n", model) %>%  
           str_to_title() %>%  
           str_replace_all("Ii", "II"), 
         meterbinnn =floor_divide(meter_max_mnth, 20000), 
         name = fct_recode(name,
           `Corrective Cost` = "event_cost_per_meter_per_mnth",
           `Total Cost` = "comb_cost",
           `Preventative Cost` = "rep_cost"
           )
         ) %>%  
  filter(`Make/Model` != "Chevrolet\nUplander" &
          `Make/Model` != "Startrans\nSenator II")

#would like to avoid making this if possible 
data2 = data1  %>% 
  group_by(meter_bin, `Make/Model`, name) %>%  
  summarise(value = median(value, na.rm = T)) %>%  
  mutate(meter_bin = as.numeric(as.character(meter_bin))) 

data2 %>% 
  filter(name == "Total Cost") %>% 
  ggplot() +
  coord_cartesian(ylim=c(0, 2)) +
  geom_boxplot(data = data1 %>%
                 filter(name != "Total Cost"), 
               aes(meter_max_mnth, value, color = name, group = interaction(name, meterbinnn, lex.order = F)),
               outlier.shape = NA) +
  geom_point(aes(meter_bin, value, color = name)) +
  geom_line(aes(meter_bin, value, color = name))  +
  facet_grid(rows = vars(`Make/Model`), scales = "free_x") +
  theme(axis.text.x = element_text(angle = 45, vjust = 0.5, hjust=.5)) + 
  labs(y = "Costs per Unit of Operating Time", 
       x = "Vehicle Mileage (20K mile bins)", 
       color = "Cost Types", 
       title = "Vehicle Operation Costs per Unit of Operation Time",
       caption = "Vertical plot range limited [0, 2] and outliers omitted for visual clarity." ) + 
  theme(legend.position="top")
```

Key takeaways (`r Sys.Date()`): 

* Key takeways from the initial iteration of this analysis still stand
  + Both replacement and maintenacne costs either become asymptotic or slowly dirft downward or upwards, respectively 
  + No definitive trough behaviopur seen in combined cost values
    - Initial assumption: combined costs would bend upwards at some point given pressure from increasing maintenace costs
  + Vehicles do exhibit an increase in combined cost at extreme ranges 
    - Likely, due to a small number of maintance events at these range bins 
  + A few vehicles show a slight but noticable increase in comvined costs
    - Uplander ~ 150K
    - Entervan ~ 150K
    - Express ~ 150K
    - GC 2 ~ 225K


