---
title: "Spurious Data Entires"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  cache = FALSE, cache.lazy = FALSE, autodep = TRUE, warning = FALSE, 
  message = FALSE, echo = TRUE, dpi = 180,
  fig.width = 8, fig.height = 5
  )
```

```{r include=FALSE}
#sources local global file
source(here::here("code/zz_localG.R")) 

#loads the data
o_wo = readRDS(here("data/o_wo.rds"))
o_wo_agg = readRDS(here("data/o_wo_agg.rds"))
udcm = readRDS(here("data/udcm.rds"))
joined = readRDS(here("data/joined.rds"))
```

## Data Issues

This section was developed to highlight potential issues with the data used for the KCM ULB Analysis. 

When you click the **Knit** button a document will be generated that includes both content as well as the output of any embedded R code chunks within the document. You can embed an R code chunk like this:


The below table is a detailed glance at the work order dataset. It depicts summary statisitics of numeric values found in or built from the original work order dataset. There are three points 

* The *P0* value for row 10, *meter_diff_mnth*, is ngative 
  + This should never be the case as that variable measures the month-to-month increase of each vehicles meter mileage - no meter can decrease with repsect to the previous month meter value
  + It appears that this issue is caused by erroneous data recording
* The *P100* values for rows 4-8, *cost tracking variables*, are all extrememly large comparaed to the rest of the individual metric distributions
  + While a fluid related maintenace cost ,*act_fluid_cost_do*, could potentially cost $558k, this value seems spurious given the rest of the data

```{r steup, include=FALSE}
o_wo_skimmed = o_wo %>%  
  skim()

o_wo_skimmed %>%  
  yank("numeric") %>%  
  dt_common()
```

The plot below is an example of the former point listed above regarding negative monthly meter difference values. Unit 577 shows a massive dip in montly cummlative meter mileage in early 2010 (bottom plot). The top plot depicts the monthly meter difference - the dip in the bottom plot is depicted as a dip and spike in the difference plot. A singular incorrect point in the cummlative vehcile miles plot manifests itself a two massivie monhtly deviations - this is not optimal as monthly meter difference is integral in calculating a vehicle's operational cost (maintenance cost per miles driven per month). 
```{r}
bad_mileage = o_wo %>%  
  pivot_longer(cols = c(meter_sans0, meter_diff_mnth)) %>%
  group_by(name) %>% 
  mutate(anomaly = case_when(abs(value-lag(value))>10000~"True", 
                             T~"False"),
         name = fct_relevel(name, "meter_diff_mnth", "meter_sans0")) %>% 
  filter(unit_id == 577) %>%  
  ggplot(aes(fisc_pd_corr, value)) + 
  geom_line() + 
  facet_grid(rows = vars(name), scales = 'free_y', 
             labeller = as_labeller(
               c(
                 `meter_sans0` = "Vehicle Meter",
                 `meter_diff_mnth` = "Monthly Meter Diff"
               ))) +
  scale_x_date(date_labels = "%Y", date_breaks = "1 year") +
  labs(y = "Value", x = "Year", title = "Example of Poor Meter Data Quality")

bad_mileage
```

```{r}
o_wo_agg 

joined %>%
  ggplot(aes(age_bin, event_cost_per_meter_per_mnth)) +
  geom_boxplot() +
  stat_summary(fun = mean, geom = "point", shape = 20, size = 10, color = "red") +
  # facet_grid(cols = vars(mcc), scales = "free") +
  ylim(0, 1)

```


